An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-03 EconGrowthSeminar.Rnw
Changes to the presentation before the seminar.

\begin{frame}{Energy is Essential for Economic Activity} %----------------------------
  \begin{center}What do you do that doesn't require energy?\end{center}
\end{frame}

@@

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Growth_Models.pdf}
  \end{center}
  \begin{itemize}
    \item{Goal: predict economic output over time}
    \item{All factors and outputs indexed to initial year: $y = Y/Y_0, e = E/E_0$, etc.}
    \vspace{\baselineskip}
    \pause

    \item{Set aside debates about growth and its measure}
  \end{itemize}
\end{frame}

@@

    Extra credit: reflect dynamics of energy cost share and price spikes.
  \end{center}
\end{frame}

@@

\begin{frame}{Coordinates of Analysis} %----------------------------
  \begin{itemize}
    \item{Type of Economy}
    \vspace{\baselineskip}
    \item{Type of Energy}
    \vspace{\baselineskip}
    \item{Economic Growth Model}
  \end{itemize}
\end{frame}

\begin{frame}{Type of Economy} %----------------------------
  \begin{center}
    \begin{tabular}{r c c l}
      \multirow{3}{*}{Developed}  & \includegraphics[height=0.11in]{flag/flag_US.png} & US & USA           \\
                                  & \includegraphics[height=0.11in]{flag/flag_UK.png} & UK & United Kingdom\\
                                  & \includegraphics[height=0.11in]{flag/flag_JP.png} & JP & Japan         \\
      \hline
      \multirow{2}{*}{BRICS}      & \includegraphics[height=0.11in]{flag/flag_CN.png} & CN & China         \\
                                  & \includegraphics[height=0.11in]{flag/flag_ZA.png} & ZA & South Africa  \\
      \hline
      \multirow{2}{*}{Petrol}     & \includegraphics[height=0.11in]{flag/flag_SA.png} & SA & Saudi Arabia  \\
                                  & \includegraphics[height=0.11in]{flag/flag_IR.png} & IR & Iran          \\
      \hline
      \multirow{2}{*}{LDCs}       & \includegraphics[height=0.11in]{flag/flag_TZ.png} & TZ & Tanzania      \\
                                  & \includegraphics[height=0.11in]{flag/flag_ZM.png} & ZM & Zambia        \\
    \end{tabular}
  \end{center}
\end{frame}

@@

\begin{frame}{Economic Growth Model} %----------------------------
  Single Factor: \sfEqn
  \vspace{\baselineskip}

@@

    \begin{tabular}{|c|l|}\hline
      \yRow\\
      \timeRow\\
      \initialTimeRow\\
      $f$ & $F$/$F_0$, factor of production ($k$, $l$, $q$, $x$, or $u$)\\
      \hline
      \lambdaRow\\
      $m$ & ``influence'' of factor $f$ ($\alpha$, $\beta$, or $\gamma$)\\
      \hline
    \end{tabular}
    \vspace{\baselineskip}

@@

@@

@@

    \item{Labor ($L$): human work in an economy (persons or hours)}
  \end{itemize}
\end{frame}

\begin{frame}{Sources of Economic Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{WB GEP} & \multirow{3}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & $Y$, $K$ for all countries & \multirow{3}{*}{Real 2005\$} \\
                              &                          & (1980-2011: Developed; & \\
                              &                          & 1991-2011: Other) & \\
      \hline
      \multirow{4}{*}{WB WDI} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & Pop, Pop < 15 yo,    & \multirow{4}{*}{Total employment (persons)} \\
                              &                          & Emp/Pop ratio > 15 yo, &                                \\
                              &                          & used to find $L$, (1991-2011: &                         \\
                              &                          & CN, ZA, SA, IR, TZ, ZM) &                         \\
      \hline
      \multirow{2}{*}{US BLS} & \multirow{2}{*}{\includegraphics[scale=0.07]{figure_other/Figure_BLS.png}} & $L$ for developed countries & Labor hours \\
                              &                          & (1980-2011: US, UK, JP)     & per year \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{Sources of Energy Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{IEA} & \multirow{3}{*}{\includegraphics[scale=0.20]{figure_other/Figure_IEA.png}} & Coal, Oil, Petroleum, & \multirow{3}{*}{GJ/year} \\
                              &                          & Natural gas, Nuclear, & \\
                              &                          & Renewables & \\
      \hline
      \multirow{2}{*}{FAOStat} & \multirow{2}{*}{\includegraphics[scale=0.06]{figure_other/Figure_FAOStat.png}} & Fuel wood,    & \multirow{2}{*}{GJ/year} \\
                              &                          & Available livestock &                                \\
      \hline
      \multirow{2}{*}{REXS} & \multirow{2}{*}{\includegraphics[scale=0.05]{figure_other/Figure_Warr.png}} & Energy allocation factors. &  \multirow{2}{*}{\citet{Warr:2006hc}}\\
                              &                          & (1980-2000: Developed)     & \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{Example: Single-factor model with $q$} %----------------------------
  \begin{center}$y = A \: q^{\gamma}; A \equiv \mathrm{e}^{\lambda (t-t_0)}$\end{center}
  \begin{itemize}
    \item{Select $\lambda$, $\gamma$ to minimize \\ $SSE_y = \sum\limits_{t=1980}^{2011}(y_{pred} - y_{act})^2$}
    \pause
    \vspace{\baselineskip}

    \item{$\lambda$ and $m$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on Single-factor Model with $q$} %----------------------------
  \begin{itemize}
    \item{Not typically used for economic modeling: illustrates confidence intervals}
    \vspace{\baselineskip}

    \item{$CI_{\gamma}$ > $CI_{\lambda}$ for all countries\footnote{\tiny as percentage of $0.0 < \lambda < 0.1$ and $0 < \gamma < 1$ \normalsize}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Two Important Questions for the\\Cobb-Douglas Model} %----------------------------
  \begin{center}
    \cdeEqn
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, if $\gamma$ not small.}
    \vspace{\baselineskip}

@@

\begin{frame}{Two Important Questions for\\CES Model} %----------------------------
  \begin{center}
    \cesEqn
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, if $\delta$ is not large.}
    \vspace{\baselineskip}

@@

\begin{frame}{$CI_{\delta}$ vs. $\delta$ for CES with $q$} %----------------------------
<<CI_delta_vs_delta_Graph, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5>>=
CIvsParamPlot(model="CES", param="delta", energyType="Q")
@
\end{frame}

@@

\begin{frame}{Parameter Estimation for LINEX Model Equation} %----------------------------
  \begin{center} \linexEqn \end{center}
  \begin{itemize}
    \item{Derived from thermodynamics \citep{Kummel:1980wx}}
    \item{Time-dependent technology $\rightarrow$ efficiency ratios}
    \item{Productivity ($A$) $\uparrow$ if}
    \begin{itemize}
      \item{Efficiency of producing capital stock increases \\ ($k$ > $l + e$)}
      \item{Energy efficiency of labor increases \\ ($l$ > $e$)}
    \end{itemize}
    \item{$a_0$ and $c_t$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Summary of Analysis} %----------------------------
  \begin{itemize}
  \item{Including energy ($E$) in Cobb-Douglas and CES production functions appears to be unwarranted}
  \pause
    \begin{itemize}
      \scriptsize
      \item{Energy parameters ($\gamma$, $\delta$) less important than others ($\lambda$, $\alpha$, $\beta$) for nearly all countries}
      \pause
      \item{Why? Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$)}
      \normalsize
    \end{itemize}
  \vspace{\baselineskip}
  \pause
  \item{Including energy ($E$) in LINEX production function is required}
  \pause
    \begin{itemize}
      \scriptsize
      \item{One factor of production: energy}
      \pause
      \item{$a_0$ and $c_t$ are significant for most countries}
      \pause
      \item{Unfortunately, LINEX fits are poorer quality than CD or CES}
      \normalsize
    \end{itemize}
  \end{itemize}
\end{frame}

@@

    Answer: Cobb-Douglas or CES without energy
  \end{center}
\end{frame}

@@

    \item{Models (CD and CES without energy) could be \emph{included within} a model that endogenizes price and cost share}
    \pause
    \item{Still searching for a modeling approach that captures price and cost share dynamics \\ \scriptsize System dynamics modeling? \normalsize}
  \end{itemize}
\end{frame}

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Verify and improve estimates of statistical properties for \emph{all} fitting parameters}
    \vspace{\baselineskip}
    \item{Incorporate all constraints for Cobb-Douglas model}
    \vspace{\baselineskip}
    \item{Submit paper}
  \end{itemize}  
\end{frame}

@@

>>>>

\begin{frame} %----------------------------
  \begin{center}What do you do that doesn't require energy?\end{center}
\end{frame}

@@

\begin{frame}{Economic (Growth) Modeling} %----------------------------
  \begin{center}
    \includegraphics[height=2.5cm]{figure_other/Growth_Models.pdf}
  \end{center}
  \begin{itemize}
    \item{Goal: predict economic output over time}
    \item{Set aside debates about growth and its measure}
    \item{All factors and outputs indexed to initial year: $y = Y/Y_0, e = E/E_0$, etc.}
    \vspace{\baselineskip}
    \pause

    \item{Uncritical use of energy-augmented growth models in the literature}
  \end{itemize}
\end{frame}

\begin{frame}{What's Needed?} %----------------------------
  \begin{itemize}
    \item{Go beyond developed economies}
    \pause
    \vspace{\baselineskip}

    \item{Rigorous, statistical evaluation of energy-augmented growth models with $Q$, $X$, and $U$}
    \pause
    \vspace{\baselineskip}

    \item{Comparative analysis of growth models}
  \end{itemize}
\end{frame}

@@

    Extra credit: reflect dynamics of energy cost share and price spikes
  \end{center}
\end{frame}

@@

\begin{frame}{Coordinates of Analysis} %----------------------------
  \begin{itemize}
    \item{Type of economy}
    \vspace{\baselineskip}
    \item{Type of energy}
    \vspace{\baselineskip}
    \item{Type of economic growth model}
  \end{itemize}
\end{frame}

\begin{frame}{Type of Economy} %----------------------------
  \begin{center}
    \begin{tabular}{r c c l}
      \multirow{3}{*}{Developed}  & \includegraphics[height=0.11in]{flag/flag_US.png} & US & USA           \\
                                  & \includegraphics[height=0.11in]{flag/flag_UK.png} & UK & United Kingdom\\
                                  & \includegraphics[height=0.11in]{flag/flag_JP.png} & JP & Japan         \\
      \hline
      \multirow{2}{*}{BRICS}      & \includegraphics[height=0.11in]{flag/flag_CN.png} & CN & China         \\
                                  & \includegraphics[height=0.11in]{flag/flag_ZA.png} & ZA & South Africa  \\
      \hline
      \multirow{2}{*}{Petrol}     & \includegraphics[height=0.11in]{flag/flag_SA.png} & SA & Saudi Arabia  \\
                                  & \includegraphics[height=0.11in]{flag/flag_IR.png} & IR & Iran          \\
      \hline
      \multirow{2}{*}{LDC}       & \includegraphics[height=0.11in]{flag/flag_TZ.png} & TZ & Tanzania      \\
                                  & \includegraphics[height=0.11in]{flag/flag_ZM.png} & ZM & Zambia        \\
    \end{tabular}
  \end{center}
\end{frame}

@@

\begin{frame}{Type of Economic Growth Model} %----------------------------
  Single Factor: \sfEqn
  \vspace{\baselineskip}

@@

    \begin{tabular}{|c|l|}\hline
      \yRow\\
      $A$ & productivity of factors of production\\
      \timeRow\\
      \initialTimeRow\\
      $f$ & $F$/$F_0$, factor of production ($k$, $l$, $q$, $x$, or $u$)\\
      \hline
      \lambdaRow\\
      $m$ & ``influence'' of factor $f$ ($\alpha$, $\beta$, or $\gamma$)\\
      \hline
    \end{tabular}
    \vspace{\baselineskip}

@@

\begin{frame}{Cobb-Douglas} %----------------------------
  \begin{center}
    \cdEqn\\
    $\alpha + \beta = 1$\\
    $0 \leq \alpha, \beta \leq 1$
    \vspace{\baselineskip}

    \begin{tabular}{|c|l|}\hline
      \lambdaRow\\
      \alphaRow\\
      \betaRow\\
      \hline
    \end{tabular}
    \vspace{\baselineskip}

@@

\begin{frame}{LINEX}
    \begin{center} \linexEqn \end{center}
    \vspace{\baselineskip}

  \begin{itemize}
    \item{Derived from thermodynamics \citep{Kummel:1980wx}}
    \item{Productivity ($A$) based on efficiency ratios}
    \item{$A$ $\uparrow$ if}
    \begin{itemize}
      \item{Efficiency of producing capital stock increases \\ ($k$ > $l + e$)}
      \item{Energy efficiency of labor increases \\ ($l$ > $e$)}
    \end{itemize}
  \end{itemize}
\end{frame}

@@

    \item{Labor ($L$): human work in an economy (persons or work hours)}
  \end{itemize}
\end{frame}

\begin{frame}{Sources of Economic Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{WB GEP} & \multirow{3}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & $Y$, $K$ for all countries & \multirow{3}{*}{Real 2005\$} \\
                              &                          & (1980-2011: Developed; & \\
                              &                          & 1991-2011: Other) & \\
      \hline
      \multirow{4}{*}{WB WDI} & \multirow{4}{*}{\includegraphics[scale=0.07]{figure_other/Figure_World_Bank.png}} & Pop, Pop < 15 yo,    & \multirow{4}{*}{Total employment (persons)} \\
                              &                          & Emp/Pop ratio > 15 yo, &                                \\
                              &                          & used to find $L$, (1991-2011: &                         \\
                              &                          & CN, ZA, SA, IR, TZ, ZM) &                         \\
      \hline
      \multirow{2}{*}{US BLS} & \multirow{2}{*}{\includegraphics[scale=0.07]{figure_other/Figure_BLS.png}} & $L$ for developed countries & Work hours \\
                              &                          & (1980-2011: US, UK, JP)     & per year \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{Sources of Energy Data} %----------------------------
  \begin{center}
    \scriptsize
    \begin{tabular}{r c c l}
      \multirow{3}{*}{IEA} & \multirow{3}{*}{\includegraphics[scale=0.20]{figure_other/Figure_IEA.png}} & Coal, Oil, Petroleum, & \multirow{3}{*}{GJ/year} \\
                              &                          & Natural gas, Nuclear, & \\
                              &                          & Renewables & \\
      \hline
      \multirow{2}{*}{FAOStat} & \multirow{2}{*}{\includegraphics[scale=0.06]{figure_other/Figure_FAOStat.png}} & Fuel wood,    & \multirow{2}{*}{GJ/year} \\
                              &                          & Available livestock &                                \\
      \hline
      \multirow{2}{*}{REXS} & \multirow{2}{*}{\includegraphics[scale=0.05]{figure_other/Figure_Warr.png}} & Energy allocation factors  &  \multirow{2}{*}{\citet{Warr:2006hc}}\\
                              &                          & (1980-2000: Developed)     & \\ 
    \end{tabular}
    \normalsize
  \end{center}
\end{frame}

@@

\begin{frame}{Example: Single-factor model with $q$} %----------------------------
  \begin{center}$y = A \: q^{\gamma}$; \small $A \equiv \mathrm{e}^{\lambda (t-t_0)}$ \normalsize \end{center}
  \begin{itemize}
    \item{Select $\lambda$, $\gamma$ to minimize \\ $SSE_y = \sum\limits_{t=1980}^{2011}(y_{pred} - y_{act})^2$}
    \pause
    \vspace{\baselineskip}

    \item{$\lambda$ and $\gamma$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Comments on Single-factor Model with $q$} %----------------------------
  \begin{itemize}
    \item{Single-factor not typically used for economic modeling: illustrates confidence intervals}
    \vspace{\baselineskip}

    \item{$CI_{\gamma}$ > $CI_{\lambda}$ for all countries\footnote{\tiny as percentage of 0.0 1/year < $\lambda$ < 0.1 1/year and $0 < \gamma < 1$ \normalsize}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Two Important Questions for the\\Cobb-Douglas Model} %----------------------------
  \begin{center}
    \cdeEqn
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, if $\gamma$ not near 0.0.}
    \vspace{\baselineskip}

@@

\begin{frame}{Two Important Questions for\\CES Model} %----------------------------
  \begin{center}
    \cesEqn
  \end{center}
  \begin{itemize}
    \item{Is energy important for this economy?\\Yes, if $\delta$ is not near 1.0.}
    \vspace{\baselineskip}

@@

\begin{frame}{$CI_{\delta}$ vs. $\delta$ for CES with $q$} %----------------------------
<<CI_delta_vs_delta_Graph_Q, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5>>=
CIvsParamPlot(model="CES", param="delta", energyType="Q")
@
\end{frame}

@@

\begin{frame}{Parameter Estimation for LINEX Model Equation} %----------------------------
  \begin{center} \linexEqn \end{center}
  \begin{itemize}
    \item{Select $a_0$ and $c_t$ to best fit historical data}
    \vspace{\baselineskip}

    \item{$a_0$ and $c_t$ estimated by \texttt{nls} function in \texttt{R}}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Summary of Analysis} %----------------------------
  \begin{itemize}
  \item{Including energy ($E$) in Cobb-Douglas and CES production functions appears to be unwarranted}
  \pause
    \begin{itemize}
      \scriptsize
      \item{Energy parameters ($\gamma$, $\delta$) less important than others ($\lambda$, $\alpha$, $\beta$) for nearly all countries}
      \pause
      \item{Why? Energy is \emph{always already} embedded in capital stock ($K$) and labor ($L$)}
      \normalsize
    \end{itemize}
  \vspace{\baselineskip}
  \pause
  \item{Including energy ($E$) in LINEX production function is required}
  \pause
    \begin{itemize}
      \scriptsize
      \item{One factor of production: energy}
      \pause
      \item{$a_0$ and $c_t$ are significant for most countries}
      \pause
      \item{Unfortunately, LINEX fits are poorer quality than CD or CES}
      \normalsize
    \end{itemize}
  \item{Uncritical use of energy-augmented growth models is problematic}
  \end{itemize}
\end{frame}

@@

    Answer: Cobb-Douglas without energy (possibly CES without energy)
  \end{center}
\end{frame}

@@

    \item{Models (CD, or CES, without energy) could be \emph{included within} a model that endogenizes price and cost share}
    \pause
    \item{Still searching for a modeling approach that captures price and cost share dynamics \\ \scriptsize System dynamics modeling? \normalsize}
  \end{itemize}
\end{frame}

\begin{frame}{Next Steps (Summer 2013)} %----------------------------
  \begin{itemize}
    \item{Evaluate CES without energy}
    \item{Verify and improve estimates of statistical properties for \emph{all} fitting parameters}
    \vspace{\baselineskip}
    \item{Incorporate all constraints for all models}
    \vspace{\baselineskip}
    \item{Submit paper}
  \end{itemize}  
\end{frame}

@@

\begin{frame}{$CI_{\delta}$ vs. $\delta$ for CES with $x$} %----------------------------
<<CI_delta_vs_delta_Graph_X, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5>>=
CIvsParamPlot(model="CES", param="delta", energyType="X")
@
\end{frame}

\begin{frame}{$CI_{\delta}$ vs. $\delta$ for CES with $u$} %----------------------------
<<CI_delta_vs_delta_Graph_U, fig.width=presentationGraph1ColWidth*1.5, fig.height=presentationGraph1RowHeight*1.5>>=
CIvsParamPlot(model="CES", param="delta", energyType="U")
@
\end{frame}
